---
title: "sequoia"
author: "Andrea Estandia"
date: "25/10/2025"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup}
# Knitr settings: 
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
options(scipen = 999)
```

```{r}
source("./src/wm-gwas_source.R")
```

Define thresholds
```{r}
SNP_CALLRATE_THRESH <- 0.90   # keep SNPs with >= 90% non-missing
IND_CALLRATE_THRESH <- 0.80   # keep individuals with >= 80% non-missing
MAF_THRESH <- 0.01            # keep SNPs with MAF > 0.01
GENOTYPE_ERROR <- 0.01        # error parameter LIBBY TO CHECK DIFFERENT VALUES from 0.0001 to 0.1
```

Read data. In this case 50000 SNPs. To get this file run generate_sequoia.sh in ARC
```{r}
snps <- GenoConvert(file.path(data_path, "processed", "wm-gwas.qc_subset.ped"))
```

Input sanitisation
```{r}
# ensure `snps` is converted to a numeric matrix and rownames are present.
# Expectation: `snps` exists in the environment: rows = individuals, cols = SNPs,
# values coded 0,1,2 (or NA).
if (!exists("snps")) stop("Object `snps` not found. Please provide genotype data in `snps`.")

# Ensure numeric matrix
GenoM <- as.matrix(snps)
mode(GenoM) <- "numeric"

# Ensure rownames exist — if not, create temporary IDs
if (is.null(rownames(GenoM)) || any(rownames(GenoM) == "")) {
  rownames(GenoM) <- paste0("IND", seq_len(nrow(GenoM)))
  warning("No rownames found in `snps`. Temporary IDs (IND1, IND2, ...) were created.")
}

```

Call-rate filtering (SNPs and individuals)
```{r}
# calculate callrates
snp_callrate <- colMeans(!is.na(GenoM))
ind_callrate <- rowMeans(!is.na(GenoM))

# report (optional)
message("Genotypes: ", nrow(GenoM), " individuals x ", ncol(GenoM), " SNPs before filtering.")
message("SNP callrate: mean=", round(mean(snp_callrate), 3),
        "  -- Individual callrate: mean=", round(mean(ind_callrate), 3))

# apply thresholds
keep_snps_idx <- which(snp_callrate >= SNP_CALLRATE_THRESH)
keep_inds_idx <- which(ind_callrate >= IND_CALLRATE_THRESH)

GenoM_qc1 <- GenoM[keep_inds_idx, keep_snps_idx, drop = FALSE]

message("After call-rate filtering: ", nrow(GenoM_qc1), " individuals x ",
        ncol(GenoM_qc1), " SNPs kept.")

```

Minimum allele frequency (MAF) filtering
```{r}
compute_maf <- function(vec) {
  vec <- vec[!is.na(vec)]
  if (length(vec) == 0L) return(NA_real_)
  # allele frequency p = sum(genotypes)/(2*N)
  p <- sum(vec) / (2 * length(vec))
  p <- min(p, 1 - p)
  as.numeric(p)
}

maf_vec <- apply(GenoM_qc1, 2, compute_maf)
keep_maf_idx <- which(!is.na(maf_vec) & maf_vec > MAF_THRESH)

GenoM_qc2 <- GenoM_qc1[, keep_maf_idx, drop = FALSE]

message("After MAF filtering (>", MAF_THRESH, "): ",
        nrow(GenoM_qc2), " individuals x ", ncol(GenoM_qc2), " SNPs kept.")
```

Construct LifeHistData for Sequoia
```{r}
IDs <- rownames(GenoM_qc2)

LifeHistData <- data.frame(
  ID = IDs,
  Sex = 3L,                # 1 = female, 2 = male, 3 = unknown
  BirthYear = as.integer(NA), # Libby to modify this so the AgePrior is the same for all individuals 
  stringsAsFactors = FALSE
)

# Optional: check for duplicated IDs (shouldn't happen)
if (any(duplicated(LifeHistData$ID))) {
  stop("Duplicated IDs found in LifeHistData: ", paste(unique(LifeHistData$ID[duplicated(LifeHistData$ID)]), collapse = ", "))
}
```

Run pedigree inference with sequoia (pedigree mode)
```{r}
SeqOUT_ped <- sequoia::sequoia(
  GenoM = GenoM_qc2,
  LifeHistData = LifeHistData,
  Module = "ped",
  Err = GENOTYPE_ERROR,
  quiet = FALSE
)
# extract pedigree object
ped <- as.data.frame(SeqOUT_ped$Pedigree, stringsAsFactors = FALSE)
head(ped)
```

Tidy pedigree and extract mother-prefix
```{r}
ped2 <- ped %>%
  mutate(
    OrigID = id,
    MotherPrefix = str_extract(id, "^[^_]+")  # everything before first underscore
  ) %>%
  select(OrigID, MotherPrefix, dam, sire)
```

Summarise by mother-prefix (families) and identify multi-sire families
```{r}
family_sire_summary <- ped2 %>%
  group_by(MotherPrefix) %>%
  summarise(
    n_offspring = n(),
    assigned_sires = list(na.omit(unique(sire))),
    n_unique_sires = length(na.omit(unique(sire))),
    sires_str = ifelse(n_unique_sires > 0, paste(na.omit(unique(sire)), collapse = ", "), NA_character_)
  ) %>%
  arrange(desc(n_unique_sires), desc(n_offspring))

# families with >1 inferred father
multi_father_families <- family_sire_summary %>% filter(n_unique_sires > 1)

# Print summaries
print(family_sire_summary)
```

Plot frequency
```{r}
hist_n_fathers <-
family_sire_summary %>%
  filter(n_unique_sires>0) %>%
  ggplot(aes(x=n_unique_sires))+
  geom_histogram()+
  theme_minimal(base_family = "roboto_condensed") +
  theme(
    panel.grid = element_blank(),                      
    axis.text.x = element_text(hjust = 1, family = "roboto_condensed"),
    axis.title = element_text(family = "roboto_condensed")
  ) + 
  labs(x="\nNumber of inferred fathers", y="\nCount", title="Number of fathers contributing to each family", subtitle="Using Sequoia and 16,000 SNPs")

ggsave(hist_n_fathers,
      filename=file.path(figures_path, "hist_n_fathers.png"),
      width=4,
      height=4)
```

Summarise by sire: how many distinct mother-prefixes does each sire appear with?
```{r}
sire_family_summary <- ped2 %>%
  filter(!is.na(sire) & sire != "") %>%
  group_by(sire) %>%
  summarise(
    n_offspring = n(),
    unique_mothers = list(sort(unique(MotherPrefix))),
    n_unique_mothers = length(unique(MotherPrefix)),
    mothers = paste(sort(unique(MotherPrefix)), collapse = ", ")
  ) %>%
  arrange(desc(n_unique_mothers), desc(n_offspring))

# sires that appear across multiple mother-prefixes
multi_family_sires <- sire_family_summary %>% filter(n_unique_mothers > 1)

# Print results
print(sire_family_summary)
```

For families with "1 sire", consider whether the data actually support monogamy vs. simply failing to detect multiple sires. 

LIBBY TO EXPLORE
```{r}

```